Letter: Performance analysis of \\ high-resolution ice sheet simulations

### 2022-06-26 paper/perfmod.tex
replace n->m to avoid Glen clash

\begin{table*}[ht]
{\normalsize
\begin{tabular}{cll}
\emph{name} & \emph{meaning} & \emph{units} \\ \hline
$\alpha$    & one fixed-geometry Stokes velocity solution requires $O(n^{1+\alpha})${\large \strut} work\\
$\beta$     & one implicit SIA geometry-update solution requires $O(n^{1+\beta})$ work \\
$\gamma$    & one implicit Stokes geometry-update and velocity solution requires $O(n^{1+\gamma})$ work \\
$D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
$L$         & width of simulation domain & km \\
$n$         & degrees of freedom: number of nodes in the horizontal mesh \\
$q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
$\Delta t$  & length of time step in model years & a \\
$U$         & representative horizontal ice velocity & $\text{km}\,\text{a}^{-1}$ \\
$\Delta x$  & representative width (diameter) of map-plane mesh cells & km
\end{tabular}
}
\caption{Parameters for performance analysis; $\alpha,\beta,\gamma,n$ are pure numbers.}
\label{tab:notation}
\end{table*}

@@

Table \ref{tab:notation} lists the parameters used in our simulation performance  analysis.  Our primary parameters are $\Delta x$, a representative value for the horizontal mesh (grid) cell diameter, and $n$, the number of nodes (vertices) in the horizontal mesh.  \emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $n\to\infty$ limits, and big-O notation is only used in this limit.  If the model domain is of width $L$ then these parameters are related by
\begin{equation}
\Delta x = O\left(\frac{L}{\sqrt{n}}\right) \quad \leftrightarrow \quad n = O\left(\frac{L^2}{\Delta x^2}\right).  \label{eq:paramrelation}
\end{equation}
That is, for a given value of $L$ there are $O(\sqrt{n})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L n^{-1})$.  However, our analysis addresses spatially-3D models with map-plane horizontal meshes.}

A numerical ice sheet model uses $n$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.  It requires $O(n)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}.  However, the \cite{IsaacStadlerGhattas2015} Stokes and \cite{BrownSmithAhmadia2013} higher-order multigrid solvers use 3D refinement, though with certain modifications regarding vertical refinement.}  The amount of fast memory needed by a running ice sheet simulation is also $O(n)$ if prior states are discarded or transferred to storage.  Ice sheet models also have $O(n)$ velocity variables, but these are not state variables because a very-viscous stress balance will compute velocity as a function of the true state variables.

@@

Regardless of time-stepping design, for each explicit time step the velocity (or velocity/pressure) solution of the stress balance equations requires effort, namely the solution of (spatial) partial differential equations.  For the Stokes model we suppose one such solution requires $O(n^{1+\alpha})$ work, which we describe the the number floating point operations in our simple performance analysis.

Of course, the power $\alpha$ depends on the design of the stress balance solver.  A direct linear algebra Stokes solver implementation might yield $\alpha=3$, or $\alpha \approx 2$ when exploiting sparsity, but a multigrid method, such as the algebraic multigrid \citep{Trottenbergetal2001} solver by \cite{IsaacStadlerGhattas2015}, reduces $\alpha$.  Their Antarctic ice sheet results show Newton-Krylov iterations \citep{Bueler2021} growing only slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[see Table 8.1]{IsaacStadlerGhattas2015}.
% see data/scalingisaac.m
A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ also close to zero for certain simplified geometries.\footnote{The \cite{IsaacStadlerGhattas2015} and \cite{BrownSmithAhmadia2013} results also show good parallel scaling, something not considered in our simplified analysis.}  At the even easier end, the isothermal SIA velocity computation is a trivialization of the Stokes problem in which the vertically-averaged velocity at each node is computed by a pointwise formula, $O(1)$ work; a velocity solution requires only $O(n)$ work ($\alpha=0$).

When analyzing solver scaling in this simple manner, one must be aware that the constant in $O(n^{1+\alpha})$ can be very large.  Furthermore it will vary greatly based on solver design.

Regardless of the stress balance source, an explicit time-stepping scheme applies the mass continuity equation to update the ice thickness using $O(n)$ work.  That is, once the velocity is computed for the geometry from the previous time step, old thickness values are replaced by new ones by a formula needing only $O(1)$ work per mesh node.

\newcommand{\oo}[1]{\displaystyle O\left(#1\right)}
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.5}
\begin{table*}[ht]
{\normalsize
\begin{tabular}{llll}
\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} & \emph{pessimistic} \\ \hline
explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, n^2}{L^2}}${\Huge \strut} \\
explicit & Stokes & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U n^{1.5+\alpha}}{L}}${\Huge \strut}\phantom{x} & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,n^{2+\alpha}}{L^2}}$ \\
implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, n^{1+\beta}}${\Huge \strut} \\
implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, n^{1+\gamma}}${\Huge \strut}
\end{tabular}
}
\caption{Asymptotic scaling of computational work, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $n\to\infty$.  See Table \ref{tab:notation} for notation.}
\label{tab:performancemodel}
\end{table*}

Now suppose that a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ time steps per model year.  Recalling the scaling of $\Delta x$ with $n$, namely that $\Delta x = O(L n^{-1/2})$, equivalently $n=O(L^2 \Delta x^{-2})$, for explicit schemes we may write the stability restriction as a required number of time-steps per model year in order to maintain stability, namely
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D n}{L^2}} \label{eq:explicitsiarequired}
\end{equation}
in the SIA and pessimistic-Stokes cases.  In the optimistic-Stokes case the estimate becomes
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U n^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
\end{equation}

@@

For unconditionally-stable implicit methods, however, $\Delta t$ does not scale with grid resolution; it has fixed value $\Delta t = 1/q$ determined by the need to resolve climatic interactions.  On the other hand, the per-step expense is much greater because nontrivial equations, indeed a free-boundary problem, must be solved.  FIXME for the unconditionally-stable implicit geometric-update scheme in the SIA $O(n^{1+\beta})$ work; \cite{Bueler2016} simplified dome geometry case has $\beta=1.8$, which is not adequate; it is not multigrid
% see data/bueler.perf and data/scalingbueler.m
the steady-state SIA solver in \cite{Bueler2016} allows $\Delta t=+\infty$, with $\Delta t = 100 \,\text{a}$ as a recovery strategy in the case of solver non-convergence

For an implicit Stokes simulation the geometry/velocity solve is assumed to be, in the absense of constraining research, $O(n^{1+\gamma})$ for some $\gamma>0$ to be determined.  The nMCD method proposed in [Bueler 2022] is expensive but multigrid, so it may turn out that $\gamma$ is small.

>>>>

\begin{table*}[ht]
{\normalsize
\begin{tabular}{cll}
\emph{name} & \emph{meaning} & \emph{units} \\ \hline
$\alpha$    & one fixed-geometry Stokes velocity solution requires $O(n^{1+\alpha})${\large \strut} work\\
$\beta$     & one implicit SIA geometry-update solution requires $O(n^{1+\beta})$ work \\
$\gamma$    & one implicit Stokes geometry-update and velocity solution requires $O(n^{1+\gamma})$ work \\
$D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
$L$         & width of simulation domain & km \\
$m$         & degrees of freedom: number of nodes in the horizontal mesh \\
$q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
$\Delta t$  & length of time step in model years & a \\
$U$         & representative horizontal ice velocity & $\text{km}\,\text{a}^{-1}$ \\
$\Delta x$  & representative width (diameter) of map-plane mesh cells & km
\end{tabular}
}
\caption{Parameters for performance analysis; $\alpha,\beta,\gamma,n$ are pure numbers.}
\label{tab:notation}
\end{table*}

@@

Table \ref{tab:notation} lists the parameters used in our simulation performance  analysis.  Our primary parameters are $\Delta x$, a representative value for the horizontal mesh (grid) cell diameter, and $m$, the number of nodes (vertices) in the horizontal mesh.  \emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits, and big-O notation is only used in this limit.  If the model domain is of width $L$ then these parameters are related by
\begin{equation}
\Delta x = O\left(\frac{L}{\sqrt{m}}\right) \quad \leftrightarrow \quad m = O\left(\frac{L^2}{\Delta x^2}\right).  \label{eq:paramrelation}
\end{equation}
That is, for a given value of $L$ there are $O(\sqrt{m})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$.  However, our analysis addresses spatially-3D models with map-plane horizontal meshes.}

A numerical ice sheet model uses $m$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.  It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}.  However, the \cite{IsaacStadlerGhattas2015} Stokes and \cite{BrownSmithAhmadia2013} higher-order multigrid solvers use 3D refinement, though with certain modifications regarding vertical refinement.}  The amount of fast memory needed by a running ice sheet simulation is also $O(m)$ if prior states are discarded or transferred to storage.  Ice sheet models also have $O(m)$ velocity variables, but these are not state variables because a very-viscous stress balance will compute velocity as a function of the true state variables.

@@

Regardless of time-stepping design, for each explicit time step the velocity (or velocity/pressure) solution of the stress balance equations requires effort, namely the solution of (spatial) partial differential equations.  For the Stokes model we suppose one such solution requires $O(m^{1+\alpha})$ work, which we describe the the number floating point operations in our simple performance analysis.

Of course, the power $\alpha$ depends on the design of the stress balance solver.  A direct linear algebra Stokes solver implementation might yield $\alpha=3$, or $\alpha \approx 2$ when exploiting sparsity, but a multigrid method, such as the algebraic multigrid \citep{Trottenbergetal2001} solver by \cite{IsaacStadlerGhattas2015}, reduces $\alpha$.  Their Antarctic ice sheet results show Newton-Krylov iterations \citep{Bueler2021} growing only slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[see Table 8.1]{IsaacStadlerGhattas2015}.
% see data/scalingisaac.m
A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ also close to zero for certain simplified geometries.\footnote{The \cite{IsaacStadlerGhattas2015} and \cite{BrownSmithAhmadia2013} results also show good parallel scaling, something not considered in our simplified analysis.}  At the even easier end, the isothermal SIA velocity computation is a trivialization of the Stokes problem in which the vertically-averaged velocity at each node is computed by a pointwise formula, $O(1)$ work; a velocity solution requires only $O(m)$ work ($\alpha=0$).

When analyzing solver scaling in this simple manner, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large.  Furthermore it will vary greatly based on solver design.

Regardless of the stress balance source, an explicit time-stepping scheme applies the mass continuity equation to update the ice thickness using $O(m)$ work.  That is, once the velocity is computed for the geometry from the previous time step, old thickness values are replaced by new ones by a formula needing only $O(1)$ work per mesh node.

\newcommand{\oo}[1]{\displaystyle O\left(#1\right)}
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.5}
\begin{table*}[ht]
{\normalsize
\begin{tabular}{llll}
\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} & \emph{pessimistic} \\ \hline
explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}${\Huge \strut} \\
explicit & Stokes & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}${\Huge \strut}\phantom{x} & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$ \\
implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, m^{1+\beta}}${\Huge \strut} \\
implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, m^{1+\gamma}}${\Huge \strut}
\end{tabular}
}
\caption{Asymptotic scaling of computational work, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $m\to\infty$.  See Table \ref{tab:notation} for notation.}
\label{tab:performancemodel}
\end{table*}

Now suppose that a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ time steps per model year.  Recalling the scaling of $\Delta x$ with $m$, namely that $\Delta x = O(L m^{-1/2})$, equivalently $m=O(L^2 \Delta x^{-2})$, for explicit schemes we may write the stability restriction as a required number of time-steps per model year in order to maintain stability, namely
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}} \label{eq:explicitsiarequired}
\end{equation}
in the SIA and pessimistic-Stokes cases.  In the optimistic-Stokes case the estimate becomes
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U m^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
\end{equation}

@@

For unconditionally-stable implicit methods, however, $\Delta t$ does not scale with grid resolution; it has fixed value $\Delta t = 1/q$ determined by the need to resolve climatic interactions.  On the other hand, the per-step expense is much greater because nontrivial equations, indeed a free-boundary problem, must be solved.  FIXME for the unconditionally-stable implicit geometric-update scheme in the SIA $O(m^{1+\beta})$ work; \cite{Bueler2016} simplified dome geometry case has $\beta=1.8$, which is not adequate; it is not multigrid
% see data/bueler.perf and data/scalingbueler.m
the steady-state SIA solver in \cite{Bueler2016} allows $\Delta t=+\infty$, with $\Delta t = 100 \,\text{a}$ as a recovery strategy in the case of solver non-convergence

For an implicit Stokes simulation the geometry/velocity solve is assumed to be, in the absense of constraining research, $O(m^{1+\gamma})$ for some $\gamma>0$ to be determined.  The nMCD method proposed in [Bueler 2022] is expensive but multigrid, so it may turn out that $\gamma$ is small.
